---
title: "Appendix 2: Benchmark and posteriros"
author: "Sebastian Sosa"
date: "2025-12-18"
output: html_document
---

# 1. Dependencies
```{r setup, include=FALSE}
#devtools::install_github('https://github.com/ctross/STRAND.git')
#devtools::install_github('https://github.com/BGN-for-ASNA/BIR')
library(BayesianInference)
library(reticulate)
library(posterior)
library(STRAND)
library(ggplot2)
library(ggrepel)
library(jsonlite)
library(tidyr)
library(dplyr)
set.seed(1)
m = importBI() # Import BI object
jnp = reticulate::import('jax.numpy') # import jax.numpy
```

# 2. Utilities functions
## Functions for posterior comparison
```{r setup, include=FALSE}
to_r = function(array){np = reticulate::import("numpy"); reticulate::py_to_r(np$array(array))}
```

```{r setup, include=FALSE}
plot_benchmark <- function(cpu, gpu){
  colnames(gpu) = c("N", "BI_gpu")
  all = merge(cpu, gpu, by = 'N')
  
  time_to_seconds <- function(time_str) {
    # Remove the negative sign and split into components
    is_negative <- grepl("^-", time_str)
    time_str <- gsub("^-", "", time_str)
    parts <- strsplit(time_str, ":")[[1]]
    hours <- as.numeric(parts[1])
    mins <- as.numeric(parts[2])
    secs <- as.numeric(parts[3])
    total_seconds <- hours * 3600 + mins * 60 + secs
    if (is_negative) total_seconds <- -total_seconds
    return(total_seconds)
  }
  all$STAN <- sapply(all$stan, time_to_seconds)
  all["BI CPU"] <- sapply(all$BI, time_to_seconds)
  all["BI GPU"]<- all$BI_gpu
  
  backend = c('STAN', 'BI CPU', 'BI GPU')
  res = NULL
  for(a in 1:length(backend)){
    res = rbind(
      res,
      data.frame("N" = all$N, "Backend" = backend[a], "Time" = 
                   abs(all[,colnames(all) %in% backend[a]])
      )
    )
  }
  
  
  first_pts <- res %>%
    group_by(Backend) %>%
    slice_min(N, n = 1)
  
  last_pts <- res %>%
    group_by(Backend) %>%
    slice_max(N, n = 1)
  
  
  p <- ggplot(res, aes(N, Time, color = Backend, group = Backend)) +
    geom_line(linewidth = 0.8, linetype = 'dashed') +
    geom_point(size = 3) +
    
    # labels at first N (show log10(Time))
    geom_text_repel(
      data = first_pts,
      aes(label = round(log10(Time), 2)),  # log scale for text
      nudge_x = -20,
      direction = "y",
      hjust = 1,
      show.legend = FALSE
    ) +
    
    # labels at last N (show log10(Time))
    geom_text_repel(
      data = last_pts,
      aes(label = round(log10(Time), 2)),  # log scale for text
      nudge_x = 20,
      direction = "y",
      hjust = 0,
      show.legend = FALSE
    ) +
    
    scale_color_manual(
      values = c(
        "STAN"   = "#D55E00",
        "BI GPU" = "#2500D6",
        "BI CPU" = "#02D600"
      )
    ) +
    
    # log10 scale for Y-axis
    scale_y_log10(
      name = "log10(Time in seconds)",    # change Y-axis label
      breaks = scales::trans_breaks("log10", function(x) 10^x),
      labels = scales::trans_format("log10", scales::math_format(10^.x))
    ) +
    
    labs(
      x = "Number of nodes",
      color = "Backend"
    ) +
    
    theme_minimal(base_size = 13) +
    theme(
      legend.position = "top",
      panel.grid.minor = element_blank(),
      axis.title = element_text(face = "bold")
    )
  ggsave("Benchmark_plot.png", plot = p, width = 8, height = 8, dpi = 300)
  return(p)
  
}
```

```{r setup, include=FALSE}
posteriors_comparison <- function(m, res_stan){
  samps = m$posteriors
  sr_sigma = to_r(samps$sr_sigma)
  sr_L = to_r(samps$sr_L)
  sr_raw = aperm(to_r(samps$sr_raw), perm = c(1, 3, 2))
  dr_L = to_r(samps$dr_L)
  dr_raw = NULL
  dr_sigma = matrix(to_r(samps$dr_sigma), ncol = 1)
  
  
  B = list()
  count = 1
  for(i in 1:length(names(samps))){
    if(grepl("b_", names(samps)[i])){
      B[[count]] = to_r(samps[[i]])
      count = count + 1 
    }
  }
  
  BI_samples = list(block_parameters = B, focal_target_sd = sr_sigma, 
                    focal_target_L = sr_L, focal_target_random_effects = sr_raw, 
                    dyadic_sd = dr_sigma, dyadic_L = dr_L, dyadic_random_effects = dr_raw, 
                    error_sd = NULL,
                    focal_effects = to_r(m$posteriors$sender_effects),
                    target_effects = to_r(m$posteriors$receiver_effects),
                    dyadic_coeffs = to_r(m$posteriors$dyad_effects))
  names(BI_samples)
  
  res = list()
  for(a in 1:length(BI_samples)){
    if(is.null( BI_samples[[a]])){next}
    if(names(BI_samples)[a] == "focal_target_random_effects"){next}
    if(is.null(dim(BI_samples[[a]]))){
      for(b in 1:length(BI_samples[[a]])){
        BI_mean = apply(BI_samples[[a]][[b]], 2, mean)
        BI_std = apply(BI_samples[[a]][[b]], 2, sd)
        
        STAN_mean = apply(res_stan$samples$srm_model_samples[[a]][[b]], 2, mean)
        STAN_std = apply(res_stan$samples$srm_model_samples[[a]][[b]], 2, sd)
        
        name = paste0(names(BI_samples)[a],'_',b)
        names = NULL
        for(c in 1:length(BI_mean)){
          names = c(names, paste0(names(BI_samples)[a],'_',b,'_',c))
        }
        
        df = data.frame(BI_mean, 
                        BI_std, 
                        STAN_mean, 
                        STAN_std, 
                        var = names)
        res[[name]] = df
      }
    }else{
      BI_mean = apply(BI_samples[[a]], 2, mean)
      BI_std = apply(BI_samples[[a]], 2, sd)
      
      STAN_mean = apply(res_stan$samples$srm_model_samples[[a]], 2, mean)
      STAN_std = apply(res_stan$samples$srm_model_samples[[a]], 2, sd)
      
      names = NULL
      for(b in 1:length(BI_mean)){
        names = c(names, paste0(names(BI_samples)[a],'_',b))
      }
      
      df = data.frame(BI_mean, BI_std, STAN_mean, STAN_std, var = names)
      res[[names(BI_samples)[a]]] = df
    }
  }
  
  library(tidyr)
  library(dplyr)
  combined_df <- bind_rows(res)
  long_df <- combined_df %>%
    pivot_longer(
      cols = c(BI_mean, STAN_mean, BI_std, STAN_std),
      names_to = c("metric", ".value"),
      names_pattern = "(.*)_(.*)"
    ) %>%
    mutate(group = paste0(var, "_", var))
  
  library(ggplot2)
  
  # Create the plot
  ggplot(long_df, aes(x = mean, y = var, group = metric, color = metric)) +
    geom_point(
      size = 3,
      position = position_nudge(y = ifelse(long_df$metric == "STAN", 0.2, 0))
    ) +
    geom_linerange(
      aes(xmin = mean - std, xmax = mean + std),
      position = position_nudge(y = ifelse(long_df$metric == "STAN", 0.2, 0)),
      linewidth = 1
    ) +
    scale_y_discrete(
      labels = c(
        "block_parameters_1_1" = "Any", 
        "block_parameters_2_1" = "Group 1-1", 
        "block_parameters_2_2" = "Group 1-2", 
        "block_parameters_2_3" = "Group 1-3", 
        "target_effects_1"= "Receiver_mass_Effect",
        "target_effects_2"= "Receiver_age_Effect",
        "target_effects_3"= "Receiver_love_Effect",
        "target_effects_4"= "Receiver_fire_Effect",
        "focal_effects_1"= "Sender_Mass_Effect",
        "focal_effects_2"= "Sender_Age_Effect",
        "focal_effects_3"= "Sender_love_Effect",
        "focal_effects_4"= "Sender_fire_Effect",
        "focal_target_sd_1" = "Nodes_random_Effect_sd1",
        "focal_target_sd_2" = "Nodes_random_Effect_sd2",
        "dyadic_coeffs_1" = "Dyads_Kinship_Effect",
        "dyadic_coeffs_2" = "Dyads_Dominance_Effect",
        "dyadic_coeffs_3" = "Dyads_Interaction_Effect",
        "focal_target_L_1" = "Nodes_random_Effect_Chol1",
        "focal_target_L_2" = "Nodes_random_Effect_Chol2",
        "dyadic_L_1" = "Dyads_random_Effect_Chol1",
        "dyadic_L_2" = "Dyads_random_Effect_Chol2",
        "dyadic_sd_1" = "Dyads_Effect_sd"
      )
    ) +
    theme_minimal()
  
  return(list(BI_samples,res_stan$samples$srm_model_samples))
}
```

```{r setup, include=FALSE}
difftime_to_hms <- function(x, assume_units = NULL, round_seconds = TRUE) {
  # x : difftime, numeric, or character like "Time difference of -15.16133 mins"
  # assume_units : one of c("secs","mins","hours","days","weeks") when x is numeric
  # round_seconds : whether to round seconds to nearest integer
  
  # helper to compute string from total seconds
  format_from_secs <- function(total_secs) {
    sign_str <- ifelse(total_secs < 0, "-", "")
    sec <- abs(total_secs)
    sec <- if (round_seconds) round(sec) else floor(sec)
    h <- sec %/% 3600
    m <- (sec %% 3600) %/% 60
    s <- sec %% 60
    sprintf("%s%02d:%02d:%02d", sign_str, h, m, s)
  }
  
  # obtain seconds_total as numeric
  if (inherits(x, "difftime")) {
    seconds_total <- as.numeric(x, units = "secs")
    return(format_from_secs(seconds_total))
  }
  
  if (is.numeric(x)) {
    if (is.null(assume_units)) {
      stop("Numeric input requires 'assume_units' (one of 'secs','mins','hours','days','weeks').")
    }
    units <- match.arg(assume_units, c("secs","mins","hours","days","weeks"))
    factor <- c(secs = 1, mins = 60, hours = 3600, days = 86400, weeks = 604800)[units]
    seconds_total <- x * factor
    return(format_from_secs(seconds_total))
  }
  
  if (is.character(x)) {
    # try parse patterns like "Time difference of -15.16133 mins" or " -2.241148 hours"
    # capture number and unit (optional)
    m <- regmatches(x, regexec("([-+]?[0-9]*\\.?[0-9]+)\\s*([a-zA-Z]+)?", x))[[1]]
    if (length(m) >= 2) {
      num <- as.numeric(m[2])
      unit <- ifelse(length(m) >= 3 && nzchar(m[3]), tolower(m[3]), "secs")
      # normalize common unit forms
      unit <- sub("^min(s)?$", "mins", unit)
      unit <- sub("^sec(s)?$", "secs", unit)
      unit <- sub("^hr(s)?$|^h$", "hours", unit)
      unit <- sub("^hour(s)?$", "hours", unit)
      unit <- sub("^day(s)?$", "days", unit)
      unit <- sub("^week(s)?$", "weeks", unit)
      if (! unit %in% c("secs","mins","hours","days","weeks")) {
        # try singular like 'min' or plural 'mins' already handled; fallback assume secs
        unit <- "secs"
      }
      factor <- c(secs = 1, mins = 60, hours = 3600, days = 86400, weeks = 604800)[unit]
      seconds_total <- num * factor
      return(format_from_secs(seconds_total))
    } else {
      stop("Could not parse numeric value from the character input.")
    }
  }
  
  stop("Unsupported input type. Provide a difftime, numeric (plus assume_units), or a character string.")
}
```

```{r setup, include=FALSE}
time_to_seconds <- function(time_str) {
  # Remove the negative sign and split into components
  is_negative <- grepl("^-", time_str)
  time_str <- gsub("^-", "", time_str)
  parts <- strsplit(time_str, ":")[[1]]
  hours <- as.numeric(parts[1])
  mins <- as.numeric(parts[2])
  secs <- as.numeric(parts[3])
  total_seconds <- hours * 3600 + mins * 60 + secs
  if (is_negative) total_seconds <- -total_seconds
  return(total_seconds)
}
```

## Main function
```{r setup, include=FALSE}
run_benchmark <- function(N_id = 50, new = TRUE, fileName = 'Benchmarks.txt'){
  # Make data ---------------
  # Covariates
  Kinship = standardize_strand(rlkjcorr( 1 , N_id , eta=1.5 ))
  Dominant = matrix(rbern(N_id*N_id), 0.2, nrow=N_id, ncol=N_id)
  Mass = rbern(N_id, 0.4)
  Age = rnorm(N_id, 0, 1)
  Love = rbern(N_id, 0.4)
  Fire = rbern(N_id, 0.6)
  
  # Organize into list
  dyadic_preds = array(NA,c(N_id,N_id,3))
  
  dyadic_preds[,,1] = Kinship
  dyadic_preds[,,2] = Dominant
  dyadic_preds[,,3] = Kinship*Dominant
  
  # Set effect sizes
  sr_mu = c(0,0)  
  sr_sigma = c(2.2, 1.7) 
  sr_rho = -0.7
  dr_mu = 0 
  dr_sigma = 3.5
  dr_rho= 0.8
  sr_effects_1 = c(2, 1.5)
  sr_effects_2 = c(-1.4, 1.0)
  sr_effects_3 = c(1.4, -1.3)
  sr_effects_4 = c(-0.9, 0)
  dr_effects_1 = c(1.5, -1.9, 2.4)
  
  # Block structure
  group_probs_block_size = c(0.25, c(0.25, 0.25)*(1-0.25))
  
  B_1 = matrix(-8,nrow=1,ncol=1)
  B_2 = matrix(rnorm(9,0,3),nrow=3,ncol=3)
  B_3 = matrix(rnorm(4,0,3),nrow=2,ncol=2)
  
  diag(B_2) = diag(B_2) + 3.5
  diag(B_3) = diag(B_3) + 3.5
  
  B = list(B_1, B_2) #, B_3)
  
  groups_1 = rep("Any",N_id) 
  groups_2 = sample( c("Red","White","Blue") , size=N_id , replace=TRUE , prob=group_probs_block_size )
  groups_3 = sample( c("Strange", "Charm") , size=N_id , replace=TRUE , prob=c(0.5,0.5) )
  
  groups = data.frame(Intercept=as.numeric(factor(groups_1)), Merica=as.numeric(factor(groups_2)))
  groups_f = data.frame(Intercept=factor(groups_1), Merica=factor(groups_2))
  individual = data.frame(Mass = Mass, Age = Age, Love = Love, Fire = Fire)
  
  #################################################### Simulate SBM + SRM network
  G = simulate_sbm_plus_srm_network(N_id = N_id, 
                                    B = B, 
                                    V = 2,
                                    groups=groups,                  
                                    sr_mu = sr_mu,  
                                    sr_sigma = sr_sigma, 
                                    sr_rho = sr_rho,
                                    dr_mu = dr_mu,  
                                    dr_sigma = dr_sigma, 
                                    dr_rho = dr_rho,
                                    error_sigma = 1.0,
                                    outcome_mode="bernoulli", 
                                    link_mode="logit",                 
                                    individual_predictors = individual,
                                    dyadic_predictors = dyadic_preds,
                                    individual_effects = cbind(sr_effects_1, sr_effects_2, sr_effects_3, sr_effects_4),
                                    dyadic_effects = dr_effects_1
  )      
  
  
  ################################################################################################ Fit STRAND
  # Create the STRAND data object
  # Add row and colnames
  name_vec = paste("Individual", 1:N_id)
  rownames(G$network) = colnames(G$network) = name_vec
  rownames(G$samps) = colnames(G$samps) = name_vec
  rownames(Kinship) = colnames(Kinship) = name_vec
  rownames(Dominant) = colnames(Dominant) = name_vec
  rownames(groups_f) = name_vec
  rownames(individual) = name_vec
  
  model_dat = make_strand_data(outcome=list(Association = G$network),  
                               block_covariates=groups_f, 
                               individual_covariates=individual, 
                               dyadic_covariates=list(Kinship=Kinship, Dominant=Dominant),  
                               outcome_mode="bernoulli", 
                               link_mode="logit",
                               exposure=list(Association = G$samps)
  )
  
  
  
  tmp = model_dat
  attributes(tmp)$class = 'list'
  json <- toJSON(tmp, pretty = TRUE, auto_unbox = TRUE)
  write(json, file = paste0('sim_',N_id,'.json'))
  
  # STAN fit -------------
  fit_stan = fit_block_plus_social_relations_model(
    data=model_dat,
    block_regression = ~ Merica,
    focal_regression = ~ Mass + Age + Love + Fire,
    target_regression = ~ Mass + Age + Love + Fire,
    dyad_regression = ~ Kinship*Dominant,
    return_predicted_network=TRUE,
    mode="mcmc",
    mcmc_parameters = list(
      chains = 1,
      iter_warmup = 1000 ,
      iter_sampling = 1000 ,
      max_treedepth = 12 ,
      refresh = 1,
      adapt_delta = 0.96)
  )
  
  res_stan = summarize_strand_results(fit_stan)
  format(object.size(res_stan), units = "Mb")
  fit_stan[[6]]
  
  
  # BI fit---------------
  jnp = reticulate::import('jax.numpy')
  m = importBI()
  any = jnp$array(model_dat$block_predictors$Intercept -1)
  merica = jnp$array(as.integer(model_dat$block_predictors$Merica)) -1
  Kinship = jnp$array(model_dat$dyadic_predictors$Kinship)
  Dominant = jnp$array(model_dat$dyadic_predictors$Dominant)
  interaction = jnp$array(Kinship*Dominant)
  
  dyadic_predictors = jnp$stack(
    list(
      m$net$mat_to_edgl(Kinship), 
      m$net$mat_to_edgl(Dominant), 
      m$net$mat_to_edgl(interaction)
    )
  )$transpose(list(1L,2L,0L))
  
  individual_predictors = jnp$array(model_dat$individual_predictors)
  N_id = as.integer(model_dat$N_id)
  jnp$unique(any, return_index = T)
  merica_grp = as.integer(length(unique(model_dat$block_predictors$Merica)))
  merica_N_by_grp = jnp$array(as.vector(table(model_dat$block_predictors$Merica)))
  names(model_dat)
  
  m$data_on_model = list(
    network = m$net$mat_to_edgl(model_dat$outcomes[,,1]),
    any = any,
    merica = merica,
    merica_N_by_grp = merica_N_by_grp,
    individual_predictors = individual_predictors,
    dyadic_predictors =dyadic_predictors
  )
  
  
  model = function(network, any, merica, individual_predictors, dyadic_predictors,
                   merica_N_by_grp){
    
    B_intercept = m$net$block_model(any, 1L, N_id, name = 'Intercept')
    B_merica = m$net$block_model(merica, 3L, merica_N_by_grp, name ='Merica')
    
    sr = m$net$sender_receiver(individual_predictors, individual_predictors)
    dr = m$net$dyadic_effect(dyadic_predictors)
    
    bi.dist.bernoulli(logits = B_intercept + B_merica +  sr + dr, obs = network)
  }
  
  
  to_r = function(array){np = reticulate::import("numpy"); reticulate::py_to_r(np$array(array))}
  BI_start_time = Sys.time()
  m$fit(model = model, num_warmup = 1000L, num_samples = 1000L)
  BI_end_time = Sys.time()
  
  # Save Time of computation---------------
  
  if(new == TRUE){
    write(c('N;stan;Numpyro;BI;'), fileName, append = FALSE)
    write(paste(N_id, difftime_to_hms(fit_stan[[6]]),  difftime_to_hms(BI_start_time - BI_end_time),'', sep = ";"), fileName, append = TRUE)
  }else{
    write(paste(N_id, difftime_to_hms(fit_stan[[6]]), difftime_to_hms(BI_start_time - BI_end_time),'', sep = ";"), fileName, append = TRUE)
  }
  print("STAN--------------")
  print(fit_stan[[6]])
  
  print("BI--------------")
  print(BI_start_time - BI_end_time)
  
  # Save Estimations---------------
  POSTERIROS = posteriors_comparison(m, res_stan)
  BI_post = POSTERIROS[[1]]
  STAN_post = POSTERIROS[[2]]
  attributes(BI_post)$class = 'list'
  attributes(STAN_post)$class = 'list'
  json1 <- toJSON(BI_post, pretty = TRUE, auto_unbox = TRUE)
  json2 <- toJSON(STAN_post, pretty = TRUE, auto_unbox = TRUE)
  write(json1, file = paste0('post_BI_',N_id,'.json'))
  write(json2, file = paste0('post_STAN_',N_id,'.json'))
}
```

# 3. Run benchmark on CPU
Time fo computation between the different backends will be saved in 'Benchmarks.txt'.
Additionally, it also save for each test the simulated data in a json file to use the same data for the gpu computation.
Finally, it also save for each test the posterior distributions in 'paste0('post_BI_',N_id,'.json')' where N_id is the number of nodes in the network.
```{r setup, include=FALSE}
test = c(50, 100, 200, 400)
for(i in 1:length(test)){
  print(paste(i, '-----------------------------------------------------'))
  if(i == 1){
    run_benchmark(N_id = test[i], new = T)
  }else{
    run_benchmark(N_id = test[i], new = F)
  }
}
```

# 4. Run benchmark on GPU
## Dependencies
```{python}
from BI import bi, jnp
from datetime import datetime
import json
import json
import numpy as np
m = bi('cpu')
```

## Utilities functions
### Import simulated data
```{python}
with open('sim_50.json', 'r', encoding='utf-8') as file:
    d1 = json.load(file)
with open('sim_100.json', 'r', encoding='utf-8') as file:
    d2 = json.load(file)
with open('/sim_200.json', 'r', encoding='utf-8') as file:
    d3 = json.load(file)
with open('sim_400.json', 'r', encoding='utf-8') as file:
    d4 = json.load(file)
```

### Converted data to numpy arrays
```{python}
def convert_arrays_to_lists(obj):
    if isinstance(obj, dict):
        return {k: convert_arrays_to_lists(v) for k, v in obj.items()}
    elif isinstance(obj, (np.ndarray, np.generic)) or hasattr(obj, 'tolist'):
        return obj.tolist()
    elif isinstance(obj, list):
        return [convert_arrays_to_lists(item) for item in obj]
    else:
        return obj
```

### Convert BI posteriors to the same format as STAN
```{python}
def convert_data(d):
    merica = jnp.array([item['Merica'] for item in d['block_predictors']])
    merica_N_by_grp = jnp.unique(merica, return_counts=True)[1]

    # Extract values for each feature and store them as columns
    features = ['Mass', 'Age', 'Love', 'Fire']
    values = [[item[feature] for item in d['individual_predictors']] for feature in features]

    # Convert to a 2D JAX array (transpose to get rows as individuals)
    individual_predictors = jnp.array(values).T  # Transpose to get shape (n_individuals, n_features)
    Kinship = m.net.mat_to_edgl(jnp.array(d['dyadic_predictors']['Kinship']))
    Dominant = m.net.mat_to_edgl(jnp.array(d['dyadic_predictors']['Dominant']))
    Intearaction = Kinship*Dominant 

    data = dict(
    	N_id = d['N_id'],
    	any = jnp.ones((d['N_id'],)),
        merica = merica,
        merica_N_by_grp = merica_N_by_grp,
        individual_predictors = individual_predictors,
        network = m.net.mat_to_edgl(jnp.array(d['outcomes'])[:,:,0]),
        dyadic_predictors = jnp.stack([Kinship, Dominant, Intearaction]).transpose(1,2,0)   
    )

    return data
```


### Model
```{python}
def model(any, merica, individual_predictors, dyadic_predictors,network,
	     N_id, merica_N_by_grp):
    B_intercept = m.net.block_model(any, 1, N_id, name = 'Intercept')
    B_merica = m.net.block_model(merica, 3, merica_N_by_grp, name ='Merica')
    
    sr = m.net.sender_receiver(individual_predictors, individual_predictors)
    dr = m.net.dyadic_effect(dyadic_predictors)
    
    m.dist.bernoulli(logits = B_intercept + B_merica +  sr + dr, obs = network)
```

### Main function
```{python}
def main(d):
    data = convert_data(d)
    print(f'N = {data['N_id']}')
    m = bi('gpu')
    m.data_on_model = data
    start = datetime.now()
    m.fit(model)
    end = datetime.now()
    delta = end - start
    total_seconds = delta.total_seconds()

    line = f"{data['N_id']};{total_seconds}\n"    
    with open("time_results.txt", "a") as file:
        file.write(line)

    # Save to JSON
    serializable_dict = convert_arrays_to_lists(tmp.posteriors)    
    with open(f"posteriors_{data['N_id']}_GPU.json", "w") as json_file:
        json.dump(serializable_dict, json_file, indent=4)
    return m
```


## Run benchmark
Time fo computation will be saved in 'time_results.txt'.
Additionally, it also save for each test the posterior distributions in file 'posteriors_{data['N_id']}.json' where N_id is the number of nodes in the network.
```{python}
D = [d1,d2,d3,d4]
for i in range(len(D)):
    main(D[i])
```


# 5. Plot Benchmarks
```{r}
cpu = read.csv("Benchmarks.txt", sep = ";")
gpu = read.csv("time_results.txt", sep = ";", header = FALSE)
plot_benchmark(cpu, gpu)
```

# 6. Plot posteriors
```{r}
posteriors_comparison <- function(BI_samples, res_stan){
  res = list()
  for(a in 1:length(BI_samples)){
    if(is.null( BI_samples[[a]])){next}
    if(names(BI_samples)[a] == "dyadic_random_effects"){next}
    if(names(BI_samples)[a] == "error_sd"){next}
    if(names(BI_samples)[a] == "focal_target_random_effects"){next}
    if(is.null(dim(BI_samples[[a]]))){
      for(b in 1:length(BI_samples[[a]])){
        BI_mean = apply(BI_samples[[a]][[b]], 2, mean)
        BI_std = apply(BI_samples[[a]][[b]], 2, sd)
        
        STAN_mean = apply(res_stan[[a]][[b]], 2, mean)
        STAN_std = apply(res_stan[[a]][[b]], 2, sd)
        
        name = paste0(names(BI_samples)[a],'_',b)
        names = NULL
        for(c in 1:length(BI_mean)){
          names = c(names, paste0(names(BI_samples)[a],'_',b,'_',c))
        }
        
        df = data.frame(BI_mean, 
                        BI_std, 
                        STAN_mean, 
                        STAN_std, 
                        var = names)
        res[[name]] = df
      }
    }else{
      BI_mean = apply(BI_samples[[a]], 2, mean)
      BI_std = apply(BI_samples[[a]], 2, sd)
      BI_ess <- apply(BI_samples[[a]], MARGIN = 2, FUN = posterior::ess_bulk)
      
      STAN_mean = apply(res_stan[[a]], 2, mean)
      STAN_std = apply(res_stan[[a]], 2, sd)
      STAN_ess <- apply(res_stan[[a]], MARGIN = 2, FUN = posterior::ess_bulk)
      
      names = NULL
      for(b in 1:length(BI_mean)){
        names = c(names, paste0(names(BI_samples)[a],'_',b))
      }
      
      df = data.frame(BI_mean, BI_std, BI_ess, STAN_mean, STAN_std, STAN_ess, var = names)
      res[[names(BI_samples)[a]]] = df
    }
  }
  
  
  combined_df <- bind_rows(res)
  long_df <- combined_df %>%
    pivot_longer(
      cols = c(BI_mean, STAN_mean, BI_std, STAN_std, BI_ess, STAN_ess),
      names_to = c("Backend", ".value"),
      names_pattern = "(.*)_(.*)"
    ) %>%
    mutate(group = paste0(var, "_", var))
  
  
  ggplot(long_df, aes(x = var, y = ess, group = Backend, color = Backend ))+geom_jitter()
  
  # Create the plot
  ggplot(long_df, aes(x = mean, y = var, group = Backend, color = Backend)) +
    geom_point(
      size = 3,
      position = position_nudge(y = ifelse(long_df$Backend == "STAN", 0.2, 0))
    ) +
    geom_linerange(
      aes(xmin = mean - std, xmax = mean + std),
      position = position_nudge(y = ifelse(long_df$Backend == "STAN", 0.2, 0)),
      linewidth = 1
    ) +
    scale_y_discrete(
      labels = c(
        "block_parameters_1_1" = "Any", 
        "block_parameters_2_1" = "Merica 1-1", 
        "block_parameters_2_2" = "Merica 1-2", 
        "block_parameters_2_3" = "Merica 1-3", 
        "target_effects_1"= "Receiver_mass_Effect",
        "target_effects_2"= "Receiver_age_Effect",
        "target_effects_3"= "Receiver_love_Effect",
        "target_effects_4"= "Receiver_fire_Effect",
        "focal_effects_1"= "Sender_Mass_Effect",
        "focal_effects_2"= "Sender_Age_Effect",
        "focal_effects_3"= "Sender_love_Effect",
        "focal_effects_4"= "Sender_fire_Effect",
        "focal_target_sd_1" = "Nodes_random_Effect_sd1",
        "focal_target_sd_2" = "Nodes_random_Effect_sd2",
        "dyadic_coeffs_1" = "Dyads_Kinship_Effect",
        "dyadic_coeffs_2" = "Dyads_Dominance_Effect",
        "dyadic_coeffs_3" = "Dyads_Interaction_Effect",
        "focal_target_L_1" = "Nodes_random_Effect_Chol1",
        "focal_target_L_2" = "Nodes_random_Effect_Chol2",
        "dyadic_L_1" = "Dyads_random_Effect_Chol1",
        "dyadic_L_2" = "Dyads_random_Effect_Chol2",
        "dyadic_sd_1" = "Dyads_Effect_sd"
      )
    ) +
    theme_minimal()
}
```

```{r setup, include=FALSE}
BI_post = fromJSON("post_BI_50.json")
STAN_post = fromJSON("post_STAN_50.json")
posteriors_comparison(BI_post,STAN_post)
```

